---
output: github_document
bibliography: vignettes/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%"
)
```

<!-- badges: start -->
[![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License](https://img.shields.io/badge/license-GPL--3-blue.svg?style=flat)](https://www.gnu.org/licenses/gpl-3.0.html)
[![Release](https://img.shields.io/github/release/inbo/EQRfishes.svg)](https://github.com/inbo/EQRfishes/releases)
![GitHub](https://img.shields.io/github/license/inbo/EQRfishes)
[![R build status](https://github.com/inbo/EQRfishes/workflows/check%20package%20on%20main/badge.svg)](https://github.com/inbo/EQRfishes/actions)
![r-universe name](https://inbo.r-universe.dev/badges/:name?color=c04384)
![r-universe package](https://inbo.r-universe.dev/badges/EQRfishes)
[![Codecov test coverage](https://codecov.io/gh/inbo/EQRfishes/branch/main/graph/badge.svg)](https://app.codecov.io/gh/inbo/EQRfishes?branch=main)
![GitHub code size in bytes](https://img.shields.io/github/languages/code-size/inbo/EQRfishes.svg)
![GitHub repo size](https://img.shields.io/github/repo-size/inbo/EQRfishes.svg)
<!-- badges: end -->

# EQRfishes

The goal of R package `EQRfishes` is to calculate an index of biotic integrity (IBI) and ecological quality ratio (EQR) based on fish measures.
Rivers and water bodies with different characteristics are assessed with different criteria, each called an indextypology, as is described in the different publications on which the calculations are based:

```{r, echo=FALSE, message=FALSE, results='asis'}
library(dplyr)
library(readr)
suppressMessages(
  read_csv2(
    system.file("extdata/index_names_english_dutch.csv", package = "EQRfishes")
  )
) |>
  filter(!is.na(.data$reference), grepl("^@", .data$reference)) |>
  distinct(reference) |>
  pull("reference") |>
  paste(collapse = ", ") |>
  cat()
```


The calculations itself are also described into detail in `vignette("overview_calculations", package = "EQRfishes")`.

## References

<div id="refs"></div>

## Installation

To install `EQRfishes` from the [INBO universe](https://inbo.r-universe.dev/builds), start a new R session and run this code (before loading any packages):

```r
# Enable the INBO universe (not needed for INBO employees, as this is the default setting)
options(
  repos = c(
    inbo = "https://inbo.r-universe.dev", CRAN = "https://cloud.r-project.org"
  )
)
# Install the packages
install.packages("EQRfishes")
```

To install EQRfishes from GitHub, start a new R session and run this code (before loading any packages):

``` r
# install.packages("remotes")
remotes::install_github("inbo/EQRfishes")
```

## Use `EQRfishes`

The main function, `calculate_eqr()`, calculates de fish-based ecological quality ratio (EQR) based on a dataset with sampling location characteristics (`data_sample`) including the indextypology that should be used and a dataset with fish measure data for each location (`data_fish`).
The function documentation of `calculate_eqr()` explains how these datasets should be composed. 

Some examples for different indextypologies:

```{r example}
library(EQRfishes)
# freshwater river indexes (lowland IBI, upstream,...)
data_sample <- read.csv2(
  system.file("testdata/freshwater_sample.csv", package = "EQRfishes")
)
data_fish <- read.csv2(
  system.file("testdata/freshwater_fish_data.csv", package = "EQRfishes")
)
# add information on index that should be calculated
index_info <- data.frame(
  order_id = 1:12,
  sample_id = c(
    11652, 13384, 8681, 13282, 13561, 8434, 4550, 11611, 13534, 13512, 8507,
    2258
  ),
  indextypology = c(
    rep("brabeel", 6), "bron", "upstream", "vlagzalm", rep("brabeel", 2),
    "forel"
  )
)
data_sample <- data_sample |>
  inner_join(index_info, by = "sample_id")
# calculate index
calculate_eqr(data_sample, data_fish)
calculate_eqr(data_sample, data_fish, output = "metric")

# lake index
data_sample <- read.csv2(
  system.file("testdata/kallemoeie_sample.csv", package = "EQRfishes")
)
data_fish <- read.csv2(
  system.file("testdata/kallemoeie_fish_data.csv", package = "EQRfishes")
)
cluster <- data.frame(
  sample_id = c("Kallemoeie_e", "Kallemoeie_f"),
  index_cluster = "Kallemoeie"
)

calculate_eqr(data_sample, data_fish, cluster = cluster)
calculate_eqr(data_sample, data_fish, output = "metric", cluster = cluster)
```

Function `determine_indextypology()` can help which indextypology would be appropriate to use for a specific location:

```{r indextypology}
# load data
data_sample <- read.csv2(
  system.file("testdata/freshwater_sample.csv", package = "EQRfishes")
) %>%
  mutate(
    index_type_code = "ZTWA"
  )
# the following function adds an indextypology to the dataset
# the result can be used to calculate the EQR
data_sample <- determine_indextypology(data_sample)
```

